RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2013-12-28 papers/rjournal/eddelbuettel-francois-stokely.Rnw
Add a large new section showing how RProtoBuf can be used to serialize all of the example datasets that ship with R, and compare the sizes with R's built-in serialziation method.
Steal a section on applications/MapReduce from the HistogramTools
vignette.  To be improved further.

Combine each of the class slot and method tables into a single table
for each class.

Add a brief section on type coercion issues for e.g. bools and int64s.

Improve the wording in various places.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@624 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

The \texttt{Message} S4 class represents Protocol Buffer Messages and
is the core abstraction of \CRANpkg{RProtoBuf}.  The class contains
the slots \texttt{pointer} and \texttt{type} as described on the
Table~\ref{Message-class-table}.

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{Message} object of the C++ proto library. Documentation for the
\texttt{Message} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.message.html#Message} \\
\hline
\texttt{type} & Fully qualified name of the message. For example a \texttt{Person} message
has its \texttt{type} slot set to \texttt{tutorial.Person} \\
\hline
\end{tabular}
\caption{\label{Message-class-table}Description of slots for the \texttt{Message} S4 class}
\end{table}

Each \texttt{Message} contains a pointer to a \texttt{Descriptor}
which defines the schema of the data defined in the Message, as well
as a number of \texttt{FieldDescriptors} for the individual fields of
the message.  In addition to the field name extractors of
\texttt{Messages} introduced in the previous section, a complete list
of Message methods is available in Table~\ref{Message-methods-table}.

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{has} & Indicates if a message has a given field.   \\
\texttt{clone} & Creates a clone of the message \\
\texttt{isInitialized} & Indicates if a message has all its required fields set\\
\texttt{serialize} & serialize a message to a file, binary connection, or raw vector\\
\texttt{clear} & Clear one or several fields of a message, or the entire message\\
\texttt{size} & The number of elements in a message field\\
\texttt{bytesize} & The number of bytes the message would take once serialized\\
\hline
\texttt{swap} & swap elements of a repeated field of a message\\
\texttt{set} & set elements of a repeated field\\
\texttt{fetch} & fetch elements of a repeated field\\
\texttt{setExtension} & set an extension of a message\\
\texttt{getExtension} & get the value of an extension of a message\\
\texttt{add} & add elements to a repeated field \\
\hline
\texttt{str} & the R structure of the message\\
\texttt{as.character} & character representation of a message\\
\texttt{toString} & character representation of a message (same as \texttt{as.character}) \\
\texttt{as.list} & converts message to a named R list\\
\texttt{update} & updates several fields of a message at once\\
\texttt{descriptor} & get the descriptor of the message type of this message\\
\texttt{fileDescriptor} & get the file descriptor of this message's descriptor\\
\hline
\end{tabular}
\end{small}
\caption{\label{Message-methods-table}Description of methods for the \texttt{Message} S4 class}
\end{table}

@@

Message descriptors are represented in R with the
\emph{Descriptor} S4 class. The class contains
the slots \texttt{pointer} and \texttt{type} :

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{Descriptor} object of the C++ proto library. Documentation for the
\texttt{Descriptor} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\hline
\texttt{type} & Fully qualified path of the message type. \\
\hline
\end{tabular}
\caption{\label{Descriptor-class-table}Description of slots for the \texttt{Descriptor} S4 class}
\end{table}

Similarly to messages, the \verb|$| operator can be used to retrieve
descriptors that are contained in the descriptor, or invoke
pseudo-methods.  Thise can be used to extract field descriptors, enum
descriptors, or descriptors for a nested type.

@@

Table~\ref{Descriptor-methods-table} provides a complete list of the
avalailable methods for Descriptors.

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\hline
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{as.list} & return a named
list of the field, enum, and nested descriptors included in this descriptor.\\
\texttt{asMessage} & return DescriptorProto message. \\
\hline
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of methods for the \texttt{Descriptor} S4 class}
\end{table}

\subsection{field descriptors}
\label{subsec-field-descriptor}

@@

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{FieldDescriptor} C++ variable \\
\hline
\texttt{name} & Simple name of the field \\
\hline
\texttt{full\_name} & Fully qualified name of the field \\
\hline
\texttt{type} & Name of the message type where the field is declared \\
\hline
\end{tabular}
\caption{\label{FieldDescriptor-class-table}Description of slots for the \texttt{FieldDescriptor} S4 class}
\end{table}

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{as.character} & Character representation of a descriptor\\
\texttt{toString} & Character
representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & Return FieldDescriptorProto message. \\
\texttt{name} & Return the name of the field descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{is\_extension} & Return TRUE if this field is an extension.\\
\texttt{number} & Gets the declared tag number of the field.\\
\texttt{type} & Gets the type of the field.\\
\texttt{cpp\_type} & Gets the C++ type of the field.\\
\texttt{label} & Gets the label of a field (optional, required, or repeated).\\
\texttt{is\_repeated} & Return TRUE if this field is repeated.\\
\texttt{is\_required} & Return TRUE if this field is required.\\
\texttt{is\_optional} & Return TRUE if this field is optional.\\
\texttt{has\_default\_value} & Return TRUE if this field has a default value.\\
\texttt{default\_value} & Return the default value.\\
\texttt{message\_type} & Return the message type if this is a message type field.\\
\texttt{enum\_type} & Return the enum type if this is an enum type field.\\
\hline
\end{tabular}
\end{small}
\caption{\label{fielddescriptor-methods-table}Description of methods for the \texttt{FieldDescriptor} S4 class}
\end{table}

@@

\subsection{enum descriptors}
\label{subsec-enum-descriptor}

@@

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{EnumDescriptor} C++ variable \\
\hline
\texttt{name} & Simple name of the enum \\
\hline
\texttt{full\_name} & Fully qualified name of the enum \\
\hline
\texttt{type} & Name of the message type where the enum is declared \\
\hline
\end{tabular}
\caption{\label{EnumDescriptor-class-table}Description of slots for the \texttt{EnumDescriptor} S4 class}
\end{table}

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{as.list} & return a named
integer vector with the values of the enum and their names.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character
representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & return EnumDescriptorProto message. \\
\texttt{name} & Return the name of the enum descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{length} & Return the number of constants in this enum.\\
\texttt{has} & Return TRUE if this enum contains the specified named constant string.\\
\texttt{value\_count} & Return the number of constants in this enum (same as \texttt{length}).\\
\texttt{value} & Return the EnumValueDescriptor of an enum value of specified index, name, or number.\\
\hline
\end{tabular}
\end{small}
\caption{\label{enumdescriptor-methods-table}Description of methods for the \texttt{EnumDescriptor} S4 class}
\end{table}

\subsection{file descriptors}
\label{subsec-file-descriptor}

@@

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{slot} & \textbf{description} \\
\hline
\texttt{pointer} & external pointer to the \texttt{FileDescriptor} object of the C++ proto library. Documentation for the
\texttt{FileDescriptor} class is available from the protocol buffer project page:
\url{http://developers.google.com/protocol-buffers/docs/reference/cpp/google.protobuf.descriptor.html#FileDescriptor} \\
\hline
\texttt{filename} & fully qualified pathname of the \texttt{.proto} file.\\
\hline
\texttt{package} & package name defined in this \texttt{.proto} file.\\
\hline
\end{tabular}
\caption{\label{FileDescriptor-class-table}Description of slots for the \texttt{FileDescriptor} S4 class}
\end{table}

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{name} & Return the filename for this FileDescriptorProto.\\
\texttt{package} & Return the file-level package name specified in this FileDescriptorProto.\\
\texttt{as.character} & character representation of a descriptor. \\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}). \\
\texttt{asMessage} & return FileDescriptorProto message. \\
\texttt{as.list} & return named list of descriptors defined in this file descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{filedescriptor-methods-table}Description of methods for the \texttt{FileDescriptor} S4 class}
\end{table}

\subsection{enum value descriptors}
\label{subsec-enumvalue-descriptor}

@@

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{slot} & \textbf{description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{EnumValueDescriptor} C++ variable \\
\hline
\texttt{name} & simple name of the enum value \\
\hline
\texttt{full\_name} & fully qualified name of the enum value \\
\hline
\end{tabular}
\caption{\label{EnumValueDescriptor-class-table}Description of slots for the \texttt{EnumValueDescriptor} S4 class}
\end{table}

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{number} & return the number of this EnumValueDescriptor. \\
\texttt{name} & Return the name of the enum value descriptor.\\
\texttt{enum\_type} & return the EnumDescriptor type of this EnumValueDescriptor. \\
\texttt{as.character} & character representation of a descriptor. \\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}). \\
\texttt{asMessage} & return EnumValueDescriptorProto message. \\
\hline
\end{tabular}
\end{small}
\caption{\label{enumvaluedescriptor-methods-table}Description of methods for the \texttt{EnumValueDescriptor} S4 class}
\end{table}

@@

\subsection{Booleans}
Bools
Int64s.

@@

Protocol Buffers are frequently used to pass data between different
systems, however, and most other systems these days have support for
64-bit integers.  To work around this, RProtoBuf allows users to get
and set 64-bit integer types by treating them as characters.

@@

\section{Related work on IDLs (greatly expanded from what you have)}

\section{Design tradeoffs: reflection vs proto compiler (not addressed
  at all in current vignettes)}

>>>>

XXX Related work on IDLs (greatly expanded )

XXX Design tradeoffs: reflection vs proto compiler

@@

The \texttt{Message} S4 class represents Protocol Buffer Messages and
is the core abstraction of \CRANpkg{RProtoBuf}. Each \texttt{Message}
contains a pointer to a \texttt{Descriptor} which defines the schema
of the data defined in the Message, as well as a number of
\texttt{FieldDescriptors} for the individual fields of the message.  A
complete list of the slots and methods for \texttt{Messages}
is available in Table~\ref{Message-methods-table}.

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|p{10cm}}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{Message} object of the C++ proto library. Documentation for the
\texttt{Message} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.message.html#Message} \\
\hline
\texttt{type} & Fully qualified name of the message. For example a \texttt{Person} message
has its \texttt{type} slot set to \texttt{tutorial.Person} \\[.3cm]
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\texttt{has} & Indicates if a message has a given field.   \\
\texttt{clone} & Creates a clone of the message \\
\texttt{isInitialized} & Indicates if a message has all its required fields set\\
\texttt{serialize} & serialize a message to a file, binary connection, or raw vector\\
\texttt{clear} & Clear one or several fields of a message, or the entire message\\
\texttt{size} & The number of elements in a message field\\
\texttt{bytesize} & The number of bytes the message would take once serialized\\
\hline
\texttt{swap} & swap elements of a repeated field of a message\\
\texttt{set} & set elements of a repeated field\\
\texttt{fetch} & fetch elements of a repeated field\\
\texttt{setExtension} & set an extension of a message\\
\texttt{getExtension} & get the value of an extension of a message\\
\texttt{add} & add elements to a repeated field \\
\hline
\texttt{str} & the R structure of the message\\
\texttt{as.character} & character representation of a message\\
\texttt{toString} & character representation of a message (same as \texttt{as.character}) \\
\texttt{as.list} & converts message to a named R list\\
\texttt{update} & updates several fields of a message at once\\
\texttt{descriptor} & get the descriptor of the message type of this message\\
\texttt{fileDescriptor} & get the file descriptor of this message's descriptor\\
\hline
\end{tabular}
\end{small}
\caption{\label{Message-methods-table}Description of slots and methods for the \texttt{Message} S4 class}
\end{table}

@@

Message descriptors are represented in R with the \emph{Descriptor} S4
class. The class contains the slots \texttt{pointer} and
\texttt{type}.  Similarly to messages, the \verb|$| operator can be
used to retrieve descriptors that are contained in the descriptor, or
invoke pseudo-methods.

@@

Table~\ref{Descriptor-methods-table} provides a complete list of the
slots and avalailable methods for Descriptors.

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|p{10cm}}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{Descriptor} object of the C++ proto library. Documentation for the
\texttt{Descriptor} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\hline
\texttt{type} & Fully qualified path of the message type. \\[.3cm]
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\hline
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{as.list} & return a named
list of the field, enum, and nested descriptors included in this descriptor.\\
\texttt{asMessage} & return DescriptorProto message. \\
\hline
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of slots and methods for the \texttt{Descriptor} S4 class}
\end{table}

\subsection{Field Descriptors}
\label{subsec-field-descriptor}

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|p{10cm}}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{FieldDescriptor} C++ variable \\
\hline
\texttt{name} & Simple name of the field \\
\hline
\texttt{full\_name} & Fully qualified name of the field \\
\hline
\texttt{type} & Name of the message type where the field is declared \\[.3cm]
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\texttt{as.character} & Character representation of a descriptor\\
\texttt{toString} & Character
representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & Return FieldDescriptorProto message. \\
\texttt{name} & Return the name of the field descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{is\_extension} & Return TRUE if this field is an extension.\\
\texttt{number} & Gets the declared tag number of the field.\\
\texttt{type} & Gets the type of the field.\\
\texttt{cpp\_type} & Gets the C++ type of the field.\\
\texttt{label} & Gets the label of a field (optional, required, or repeated).\\
\texttt{is\_repeated} & Return TRUE if this field is repeated.\\
\texttt{is\_required} & Return TRUE if this field is required.\\
\texttt{is\_optional} & Return TRUE if this field is optional.\\
\texttt{has\_default\_value} & Return TRUE if this field has a default value.\\
\texttt{default\_value} & Return the default value.\\
\texttt{message\_type} & Return the message type if this is a message type field.\\
\texttt{enum\_type} & Return the enum type if this is an enum type field.\\
\hline
\end{tabular}
\end{small}
\caption{\label{fielddescriptor-methods-table}Description of slots and
  methods for the \texttt{FieldDescriptor} S4 class}
\end{table}

@@

\subsection{Enum Descriptors}
\label{subsec-enum-descriptor}

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|p{10cm}}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{EnumDescriptor} C++ variable \\
\hline
\texttt{name} & Simple name of the enum \\
\hline
\texttt{full\_name} & Fully qualified name of the enum \\
\hline
\texttt{type} & Name of the message type where the enum is declared \\[.3cm]
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\texttt{as.list} & return a named
integer vector with the values of the enum and their names.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character
representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & return EnumDescriptorProto message. \\
\texttt{name} & Return the name of the enum descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{length} & Return the number of constants in this enum.\\
\texttt{has} & Return TRUE if this enum contains the specified named constant string.\\
\texttt{value\_count} & Return the number of constants in this enum (same as \texttt{length}).\\
\texttt{value} & Return the EnumValueDescriptor of an enum value of specified index, name, or number.\\
\hline
\end{tabular}
\end{small}
\caption{\label{enumdescriptor-methods-table}Description of slots and methods for the \texttt{EnumDescriptor} S4 class}
\end{table}

\subsection{File Descriptors}
\label{subsec-file-descriptor}

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|p{10cm}}
\hline
\textbf{slot} & \textbf{description} \\
\hline
\texttt{pointer} & external pointer to the \texttt{FileDescriptor} object of the C++ proto library. Documentation for the
\texttt{FileDescriptor} class is available from the protocol buffer project page:
\url{http://developers.google.com/protocol-buffers/docs/reference/cpp/google.protobuf.descriptor.html#FileDescriptor} \\
\hline
\texttt{filename} & fully qualified pathname of the \texttt{.proto} file.\\
\hline
\texttt{package} & package name defined in this \texttt{.proto} file.\\[.3cm]
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\texttt{name} & Return the filename for this FileDescriptorProto.\\
\texttt{package} & Return the file-level package name specified in this FileDescriptorProto.\\
\texttt{as.character} & character representation of a descriptor. \\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}). \\
\texttt{asMessage} & return FileDescriptorProto message. \\
\texttt{as.list} & return named list of descriptors defined in this file descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{filedescriptor-methods-table}Description of slots and methods for the \texttt{FileDescriptor} S4 class}
\end{table}

\subsection{Enum Value Descriptors}
\label{subsec-enumvalue-descriptor}

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|p{10cm}}
\hline
\textbf{slot} & \textbf{description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{EnumValueDescriptor} C++ variable \\
\hline
\texttt{name} & simple name of the enum value \\
\hline
\texttt{full\_name} & fully qualified name of the enum value \\[.3cm]
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\texttt{number} & return the number of this EnumValueDescriptor. \\
\texttt{name} & Return the name of the enum value descriptor.\\
\texttt{enum\_type} & return the EnumDescriptor type of this EnumValueDescriptor. \\
\texttt{as.character} & character representation of a descriptor. \\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}). \\
\texttt{asMessage} & return EnumValueDescriptorProto message. \\
\hline
\end{tabular}
\end{small}
\caption{\label{EnumValueDescriptor-methods-table}Description of slots
  and methods for the \texttt{EnumValueDescriptor} S4 class}
\end{table}

@@

One of the benefits of using an Interface Definition Language (IDL)
like Protocol Buffers is that it provides a highly portable basic type
system that different language and hardware implementations can map to
the most appropriate type in different environments.
Table~\ref{table-get-types} details the correspondance between the
field type and the type of data that is retrieved by \verb|$| and \verb|[[|
extractors.

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{|c|p{5cm}p{5cm}|}
\hline
field type & R type (non repeated) & R type (repeated) \\
\hline
\hline
double	& \texttt{double} vector & \texttt{double} vector \\
float	& \texttt{double} vector & \texttt{double} vector \\
\hline
int32	  & \texttt{integer} vector & \texttt{integer} vector \\
uint32	  & \texttt{integer} vector & \texttt{integer} vector \\
sint32	  & \texttt{integer} vector & \texttt{integer} vector \\
fixed32	  & \texttt{integer} vector & \texttt{integer} vector \\
sfixed32  & \texttt{integer} vector & \texttt{integer} vector \\
\hline
int64	  & \texttt{integer} or \texttt{character}
vector \footnotemark & \texttt{integer} or \texttt{character} vector \\
uint64	  & \texttt{integer} or \texttt{character} vector & \texttt{integer} or \texttt{character} vector \\
sint64	  & \texttt{integer} or \texttt{character} vector & \texttt{integer} or \texttt{character} vector \\
fixed64	  & \texttt{integer} or \texttt{character} vector & \texttt{integer} or \texttt{character} vector \\
sfixed64  & \texttt{integer} or \texttt{character} vector & \texttt{integer} or \texttt{character} vector \\
\hline
bool	& \texttt{logical} vector & \texttt{logical} vector \\
\hline
string	& \texttt{character} vector & \texttt{character} vector \\
bytes	& \texttt{character} vector & \texttt{character} vector \\
\hline
enum & \texttt{integer} vector & \texttt{integer} vector \\
\hline
message & \texttt{S4} object of class \texttt{Message} & \texttt{list} of \texttt{S4} objects of class \texttt{Message} \\
\hline
\end{tabular}
\end{small}
\caption{\label{table-get-types}Correspondance between field type and
  R type retrieved by the extractors. \footnotesize{1. R lacks native
  64-bit integers, so the \texttt{RProtoBuf.int64AsString} option is
  available to return large integers as characters to avoid losing
  precision.  This option is described in Section~\ref{sec:int64}}.}
\end{table}

\subsection{Booleans}

R booleans can accept three values: \texttt{TRUE}, \texttt{FALSE}, and
\texttt{NA}.  However, most other languages, including the protocol
buffer schema, only accept \text{TRUE} or \text{FALSE}.  This means
that we simply can not store R logical vectors that include all three
possible values as booleans.  The library will refuse to store
\texttt{NA}s in protocol buffer boolean fields, and users must instead
choose another type (such as integers) capable of storing three
distinct values.

<<>>=
a <- new(protobuf_unittest.TestAllTypes)
a$optional_bool <- TRUE
a$optional_bool <- FALSE
<<eval=F>>=
a$optional_bool <- NA
<<echo=FALSE,eval=TRUE,print=TRUE>>=
try(a$optional_bool <- NA,silent=TRUE)
@

@@

Protocol Buffers are frequently used to pass data between different
systems, however, and most other modern systems do have support for
64-bit integers.  To work around this, RProtoBuf allows users to get
and set 64-bit integer types by treating them as characters.

@@

\section{Evaluation: data.frame to Protocol Buffer Serialization}

Saptarshi Guha wrote the RHIPE package \citep{rhipe} which includes
protocol buffer integration with R.  However, this implementation
takes a different approach: any R object is serialized into a message
based on a single catch-all \texttt{proto} schema.  Jeroen Ooms took a
similar approach influenced by Saptarshi in his \pkg{RProtoBufUtils}
package.  Unlike Saptarshi's package, however, RProtoBufUtils depends
on RProtoBuf for underlying message operations.  This package is
available at \url{https://github.com/jeroenooms/RProtoBufUtils}.

The \textbf{RProtoBufUtils} package by Jereoen Ooms provides a
\texttt{serialize\_pb} method to convert R objects into serialized
protocol buffers in this format, and the \texttt{can\_serialize\_pb}
method can be used to determine whether the given R object can safely
be expressed in this way.  To show how how this method works, we
attempt to convert all of the built-in datasets from R into this
serialized protocol buffer representation.

<<echo=TRUE>>=
library(RProtoBufUtils)

datasets <- subset(as.data.frame(data()$results), Package=="datasets")
datasets$load.name <- sub("\\s+.*$", "", datasets$Item)
n <- nrow(datasets)
@

There are \Sexpr{n} standard data sets included in R.  We use the
\texttt{can\_serialize\_pb} method to determine how many of those can
be safely converted to a serialized protocol buffer representation.

<<echo=TRUE>>=
datasets$valid.proto <- sapply(datasets$load.name, function(x) can_serialize_pb(eval(as.name(x))))
datasets <- subset(datasets, valid.proto==TRUE)
m <- nrow(datasets)
@

\Sexpr{m} data sets could be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%).  The next section illustrates how
many bytes were usued to store the data sets under four different
situations (1) normal R serialization, (2) R serialization followed by
gzip, (3) normal protocol buffer serialization, (4) protocol buffer
serialization followed by gzip.

\subsection{Compression Performance}
\label{sec:compression}

<<echo=FALSE,print=FALSE>>=
datasets$object.size <- unname(sapply(datasets$load.name, function(x) object.size(eval(as.name(x)))))

datasets$R.serialize.size <- unname(sapply(datasets$load.name, function(x) length(serialize(eval(as.name(x)), NULL))))

datasets$R.serialize.size <- unname(sapply(datasets$load.name, function(x) length(serialize(eval(as.name(x)), NULL))))

datasets$R.serialize.size.gz <- unname(sapply(datasets$load.name, function(x) length(memCompress(serialize(eval(as.name(x)), NULL), "gzip"))))

datasets$RProtoBuf.serialize.size <- unname(sapply(datasets$load.name, function(x) length(serialize_pb(eval(as.name(x)), NULL))))

datasets$RProtoBuf.serialize.size.gz <- unname(sapply(datasets$load.name, function(x) length(memCompress(serialize_pb(eval(as.name(x)), NULL), "gzip"))))

clean.df <- data.frame(dataset=datasets$load.name,
                       object.size=datasets$object.size,
                       "serialized"=datasets$R.serialize.size,
                       "gzipped serialized"=datasets$R.serialize.size.gz,
                       "RProtoBuf"=datasets$RProtoBuf.serialize.size,
                       "gzipped RProtoBuf"=datasets$RProtoBuf.serialize.size.gz,
                       check.names=FALSE)
@

Table~\ref{tab:compression} shows the sizes of 50 sample R datasets as
returned by object.size() compared to the serialized sizes.
The summary compression sizes are listed below, and a full table for a
sample of 50 datasets is included on the next page.  Sizes are comparable
but protocol buffers provide simple getters and setters in multiple
languages instead of requiring other programs to parse the R
serialization format \citep{serialization}.  One takeaway from this
table is that RProtoBuf does not in general provide any significant
space-savings over R's normal serialization mechanism.  The benefit
from RProtoBuf comes from its interoperability with other
environments.
